River networks dampen long-term hydrological signals of climate change

### 2017-03-15 drafts/PNAS/ms_pnas.tex
Monthly Color Gradient / Fist MS PNAS Draft
Made the monthly color gradients scale across all climate index values
rather than just within a month. Also adding PNAS draft. Also, adding a
bit of code that plays with the effect of N on SD when SD is expected
to grow with N. Finally, added a bit of code that shows how climate
index values become less strong and homogenous across the Fraser basin
during seasonal transitions.

\documentclass[linenumbers,draft]{AGUJournal}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{upgreek}
\usepackage{natbib}
\usepackage[parfill]{parskip}
\journalname{Geophysical Research Letters}
\begin{document}
\title{River networks dampen long-term hydrological signals of climate change}
\authors{K. A. Chezik\affil{1},
	S. C. Anderson\affil{2},
J. W. Moore\affil{1}}

\affiliation{1}{Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, 8888 University Dr., Burnaby, British Columbia V5A 1S6, Canada}
\affiliation{2}{School of Aquatic and Fishery Sciences, University of Washington, Box 455020, Seattle, WA 98195, USA}

\correspondingauthor{K. A. Chezik}{kchezik@sfu.ca}

\begin{keypoints}
\item River networks act as diversified portfolios of climate; by integrating varied climate signals the network substantially dampens downstream climate-driven hydrologic trends.
\item River network climate dampening varied seasonally in the Fraser River basin, declining sharply with the spring freshet, suggesting that basin-wide events homogenize the climate portfolio, thereby degrading the network dampening effect.
\item Rivers that integrate a diverse climate portfolio offer climate change relief to riverine biota that is many times greater than homogenized systems. Leveraging network dampening properties by promoting land use practices that maintain or enhance climate portfolio diversity may be a powerful tool in mitigating climate change impacts. 
\end{keypoints}

\begin{abstract}
River networks may dampen local hydrologic climate change by aggregating a diverse climate portfolio with variable upstream climate signals. Here we examine this hypothesis using flow trend estimates (1970--2007) at 55 hydrometric gauge stations within the Fraser River watershed in British Columbia, Canada. The Fraser River dampened variability in downstream long-term median discharge by 95\% from between -6 and 18\% per-decade among small watersheds to between 0.8 and 1.6\% per-decade near the confluence. A null model approach indicated that dampening of annual median flow trends within the Fraser River was typically 5.4 times greater than would be expected of a watershed exhibiting a consistent basin-wide climate response. Although climate dampening strength varied seasonally, our findings indicate that large free-flowing rivers offer a powerful and largely unappreciated process of climate change mitigation. River networks that integrate a diverse climate portfolio can dampen local extremes and offer climate relief to riverine biota.
\end{abstract}

\section{Introduction} 
As billions are spent adapting to climate change by such means as building dikes or improving stormwater drainage \citep{Narain:2011}, certain habitats and processes provide a natural defense system that mitigates climate change impacts \citep{Jones:2012}. For example, coastal habitats such as oyster reefs, mangrove forests, and eelgrass beds shield 67\% of the United States coastline and 1.4 million people from sea level rise and storms \citep{Arkema:2013}. Protection and enhancement of these natural systems has become a key component of many climate mitigation strategies \citep{Guerry:2015}, as they return multiple cost-effective services in contrast to engineered alternatives \citep{Jones:2012}. For instance, revitalization of the Mississippi delta not only improves storm protection, but also increases production and sustainability of fisheries and wildlife resources, protects the water supply, and reduces the impact of wastewater effluent \citep{Conservation:1998}. The value of climate buffering habitats stands only to increase as average global temperatures climb. Recent international agreements in Paris aim to limit global temperature rise to 1.5\degree{C} \citep{Hulme:2016}, a goal that concedes a minimum doubling of observed warming \citep{Hartmann:2013}; therefore, an urgent challenge and opportunity is discovering systems that naturally buffer against oncoming impacts of climate change.

River systems may offer an unappreciated defense against some aspects of climate change. Climate change is altering global precipitation patterns \citep{Donat:2016} and subsequently river flow regimes \citep{Hartmann:2013,Palmer:2009}. Earlier snowmelt \citep{Rauscher:2008}, reduced snow pack \citep{McCabe:2014}, shifts from snow to rain, and changes in the annual distribution of precipitation are all impacting stream flow \citep{Hartmann:2013}, resulting in flashier flows with increased frequency of flooding \citep{Hirabayashi:2013}, longer periods of drought and critically low flows for wildlife and agriculture \citep{Melillo:2014}. However, we hypothesize that large free-flowing river networks may naturally dampen these signals of climate change by integrating across different manifestations of climate. Processes that aggregate across asynchronous components tend to dampen the aggregate, a process known as the portfolio effect \citep{Doak:1998}. Climate is expressed differently and asynchronously over the landscape \citep{George:2015} and on account of their branching architecture and directional flow, river networks directionally integrate these varied climate manifestations \citep{Peterson:2013}, potentially dampening the local climatic response of sub-catchments. Large free-flowing rivers may have diverse climate portfolios, with sub-catchments acting as assets in their portfolio, thereby dampening local climate impacts. Portfolio effects in rivers are already known to dampen short-term variation \citep{Moore:2015,Yeakel:2014} but attenuation of long-term climate driven shifts have yet to be examined.

Here we examine whether rivers attenuate local long-term climate change through analyses of hydrometric data from one of the largest free-flowing rivers in North America. Located in British Columbia, Canada, the Fraser River drains an area approximately the size of the United Kingdom (\texttildelow 217,000 km\textsuperscript{2}) and in a rare combination is fairly well monitored while also having no dams on its main stem \citep{Vorosmarty:2010} (Figure~\ref{figone}). This watershed drains interior high plateau, coastal mountains, and the Canadian Rockies, and thus integrates a diverse mosaic of landscapes, weather, and climate. As with other mid-latitude rivers \citep{Bindoff:2013}, the Fraser River has expressed increasing trends in discharge volume and variability over the last several decades \citep{Dery:2012,Morrison:2002}. The river's discharge is monitored by hydrometric gauge stations throughout the watershed. We analyzed 38 years of data collected concurrently at 55 of these stations and applied a novel analytical approach to examine whether this large river exhibits signals of long-term climate dampening.

\begin{figure}[h]
	\centerline{\includegraphics[height=4in]{Fig1_Map2.pdf}}
	\caption{Flow gauge sites within the Fraser River basin (dark grey) in British Columbia Canada (light grey). Flow gauge sites (yellow dots) are scaled by the size of the contributing area. The map is projected in Albers BC UTMs to provide an equal area depiction of the region but labels are expressed in WGS84 latitude and longitude.}
	\label{figone}
\end{figure}

\section{Data and Methods}
\subsection{Discharge Data and Flow Metrics}
To assess signals of climate change in hydrology within the Fraser River basin we downloaded daily flow data from Environment Canada's HYDAT database. Within HYDAT we selected hydrometric gauge stations using three criteria: (i) the station was within the Fraser River basin, (ii) the station's flow data were not observed to be dam influenced, and (iii) the station collected data in each month between 1970 and 2007 where no month was missing more than 5 days of data and at least 35 complete years were present. We selected this 38-year range because this period maximized the number of gauge stations in the Fraser River basin (\textit{n}=55) operating concurrently over a long enough time period in which climate trends may be observable. We interpolated missing data using a linear model in log-space, but of the 763,345 days of data only 23 days were interpolated. Because dams control flow and stabilize flow trends, we removed sites where daily flow exhibited unusually high autocorrelation and low variability relative to other sites of similar size, and were observed downstream of, and atypically close to, a dam (\textit{n}=3). The downstream distance of gauge stations from dams was determined using GIS spatial layers available in the \textit{BC Data Catalog}, maintained by the provincial government of British Columbia.

As climate change alters precipitation patterns from snow to rain in temperate regions, rivers are generally predicted to have lower low-flows and higher high-flows, as well as earlier snowmelt \citep{Nijssen:2001}. These manifestations of rising temperatures are impacting species \citep{Xenopoulos:2006} and people \citep{Hirabayashi:2013}. We summarized our daily discharge data into annual and monthly flow metrics that capture these transitions. We used maximum- and minimum-flow estimates to capture trends in extreme discharge and median-flow estimates to capture general annual and seasonal trends. Because the timing of flow is particularly important to the phenology of many organisms, we also calculated the day-of-year in which half the annual flow was reached.

\subsection{River Discharge Trend Analysis}
For a river network to act as a portfolio of climate and dampen the effects of climate change, expressions of climate and their trends over time need to vary throughout the network (Figure~\ref{figtwo}). We estimated annual and monthly flow trends for each response variable at each gauge station using a generalized least squares (GLS) model with an AR1 autocorrelation function: 
\begin{linenomath*}
\begin{equation}
  \mathrm{flow}_{s,t} = a_s + b_s \mathrm{year}_{s,t} + \epsilon_{s,t}, \quad 
  \epsilon_{s,t} \sim \mathcal{N}(\upphi \epsilon_{s,t-1}, \sigma_\mathrm{flow}^2) \label{eq1},
\end{equation}
\end{linenomath*}
where $\mathrm{flow}_{s,t}$ represents a flow metric [log(maximum), log(minimum), log(median), logit(day-of-year to half-annual-flow)] at each site ($s$) and time point ($t$),  and $\mathrm{year}_{s,t}$ represents the time in years (e.g., 1970, 1971, \ldots, 2007) with 1988 subtracted to approximately center the predictor. Parameters $b_{s}$ (slope) and $a_{s}$ (intercept) represent the estimated mean effect of time on $\mathrm{flow}_{s,t}$ and the estimate of $\mathrm{flow}_{s,t}$ at time zero (i.e., 1988) respectively. Error ($\epsilon_{s,t}$) of the current time step for a given site was allowed to be correlated with that of the previous time step by $\upphi$ and was assumed to be normally distributed with a variance of $\sigma^{2}$. To ensure our model estimates and simulations remained within the calendar year we logit transformed our flow-timing response variable (i.e., day-of-year to half-annual-flow) after scaling the data between 0 and 1 (i.e., dividing by 365).

\begin{figure}[h]
	\centerline{\includegraphics[height=3in]{Fig2_Flow-Time.pdf}}
	\caption{Annual median flow at 55 sites within the Fraser River basin. Flow rates are scaled at each site by the root mean square in order to make flow rates comparable among sites. Lines are loess smoothers with a smoothing span of 0.25 predicted each month between 1970 and 2007. Line and point size are scaled to each flow gauge site's watershed area. The data show inter-annual variability within sites with varying magnitudes and directions among sites, illustrating variable long-term change in hydrology across the basin.}
	\label{figtwo}
\end{figure}

\subsection{River Network Trend Attenuation Analysis}
We hypothesized that larger catchments within the Fraser River watershed with more diverse climate portfolios would have dampened long-term hydrological change when compared to their smaller sub-catchments. Therefore, to test whether networks dampen climate signals we regressed site-specific trend estimates (e.g., \% change$\cdot$decade\textsuperscript{-1}) onto each site's catchment area using a generalized least squares model:
\begin{linenomath*}
\begin{equation}
	\hat{b}_{s} = c + d\sqrt{\mathrm{area}_{s}} + \eta_{s}, \quad
  \eta_{s} \sim \mathcal{N}(0, f(\mathrm{area}_{s})) \label{eq2},
\end{equation}
\end{linenomath*}
where $\hat{b}_{s}$ represents a flow trend at a given site ($s$) and $\mathrm{area}_{s}$ represents site $s$'s watershed area. Fitted $d$ and $c$ parameters represent the mean effect of $\sqrt{\mathrm{watershed area}}$ on flow trends and the mean basin-wide flow trend at a theoretical watershed area of zero, respectively. To quantify attenuation, we modeled the change in trend variability with the change in catchment area using an exponential variance function \citep[p.~211]{Pinheiro:2006}: 
\begin{linenomath*}
\begin{equation}
	f(\mathrm{area}_{s})=\sigma_b^2 \exp(2\updelta\sqrt{\mathrm{area}_{s}}) \label{eq3},
\end{equation}
\end{linenomath*} 
where the variance of the estimated error ($\eta$) changes exponentially with increasing $\sqrt{\mathrm{area}}$. This variance function allowed us to predict the range of flow trend values that would be expected as watershed area increased. A decreasing range of flow trends with increasing watershed area would indicate a climate-dampening network effect.

\subsection{River Discharge Trend Simulation}
In the observed data, small watersheds exhibited greater variability around their trend estimates than large watersheds, likely because of greater short-term variation in small catchments \citep{Moore:2015}. This relationship between watershed size and trend certainty may pull small watershed trend estimates away from zero, thereby creating the appearance of decreasing trend variability among sites as watershed size increases. Using a null-hypothesis framework we simulated time-series for each site with no underlying trend. These simulations were generated by a random walk process with the observed standard deviation and autocorrelation parameters ($\hat{\upsigma}$, $\hat{\upphi}$) at our sites. Using the same GLS model and AR1 correlation structure as applied to the observed data (eq.~\ref{eq1}), we estimated trends for 1000 simulated time series at each site for each response variable (e.g., Figure S3). This simulation process, a form of parametric bootstrapping, created distributions of null-expectations with which we compared our observed results.

\begin{figure}[h]
	\centerline{\includegraphics[height=4.5in]{Fig3_Annual-Funnel.pdf}}
	\caption{Annual flow trend attenuation within the Fraser River basin. (\textbf{Left}) Trend estimates $\pm$ one standard error (SE, grey) plotted against watershed area (km\textsuperscript{2}). Blue represents observed attenuation; orange and red represent simulated attenuation that is weaker and stronger than observed. (\textbf{Right}) Density plots show null model simulated variance exponents ($\hat{\updelta}$) and the proportion on either side of the observed variance exponent (blue). Flow metrics include long-term flow-timing shifts (change per decade in day-of-year (DOY) to half annual flow), where decreasing trends suggest more annual flow is occurring earlier in the year (a), and the percent change per decade in minimum (b), maximum (c) and median (d) annual flow. Simulated lines ignore variance in the intercept and slope to focus visually on attenuation.}
	\label{figthree}
\end{figure}

\subsection{Dampening Summary Statistics}
We calculated attenuation certainty as the percentage of simulated $\hat{\updelta}$ that were less than the observed $\hat{\updelta}$. We calculated attenuation strength as the ratio between the standard deviation at the smallest and largest watersheds as defined by the exponential variance function (i.e.,\\$\sqrt{f(\mathrm{area}_1)} / \sqrt{f(\mathrm{area}_2)}$ ). To compare the degree of attenuation observed with the null-simulated attenuation, we compared this ratio with the same ratio calculated from the null-simulated data (i.e., $\left( \sqrt{f(\mathrm{area}_1)} / \sqrt{f(\mathrm{area}_2)} \right) / \left( \sqrt{f(\mathrm{area}_{\mathrm{null} 1})} / \sqrt{f(\mathrm{area}_{\mathrm{null} 2})} \right)$ ). Finally, to estimate how the day-of-year to half-annual-flow was changing, we estimated the steepest point of each site's logistic curve and multiplied this slope value by 365 to revert from our 0--1 scale to our original annual range of 0--365. To get a decadal rate we multiplied these results by 3.8 (number of decades in 38 years). We estimated the steepest part of the logistic curve using the \textit{divide by four} rule \citep{Gelman:2008}, where dividing the coefficient by four equals the first derivative (slope) of the logistic curve at its steepest point.

\section{Results}
\subsection{Hydrologic Trend Diversity}
Long-term changes in flow metrics varied substantially across the watershed (Figure~\ref{figthree}). For instance, annual maximum flow decreased in 38 sites but increased in 17 over the last four decades (-28 to 15\% change$\cdot$decade\textsuperscript{-1}, std. dev. = 6\%). Annual median flow showed similar variability but a greater tendency towards rising flow trends with 49 increasing and only six decreasing (-5 to 25\% change$\cdot$decade\textsuperscript{-1} std. dev. = 5\%). Thus, within the Fraser River basin, there are a variety of long-term trends in flow, suggesting a diversity of climate responses.

\subsection{Annual Hydrologic Trend Dampening}
Long-term changes in hydrology were less variable in larger catchments, as predicted, demonstrating between 85 and 92\% dampening of flow trends moving from headwater catchments to the network's confluence. In other words, the largest catchments were approximately 10 times more stable in their climate response than the smallest catchments. For instance, trends ranged between a 6\% reduction and 19\% increase in median flow per decade among small watersheds, while large watersheds ranged between 0.8 and 1.6\% change per decade. Similar attenuation was seen in flow timing with a 94\% reduction in the day-of-year to half-annual-flow trends such that small watersheds were reaching half their annual flow between 8 days later and 25 days earlier while large watersheds have only shifted 2 to 4 days earlier between 1970 and 2007.

A null model approach illustrated that support for a river network portfolio effect was greater than 90\% for three of four annual flow trend metrics and as high as 98\% for median-annual-flow trends (Figure~\ref{figthree}). Trend attenuation was 3.1 (0.7--9.1 this and hereafter are 90\% intervals), 1.6 (0.3--4.6), 3.4 (1.0--8.5) and 5.4(1.5--14.6) times greater than the null model for annual-flow timing, and minimum, maximum and median annual-flow respectively. These statistics provide strong support for the hypothesis that river network portfolio effects contribute to climate dampening in large rivers.

\subsection{Seasonal Hydrologic Trend Dampening}
Regional flow-trends and attenuation strength varied by season. Analysis of monthly trends revealed that winter flows have been getting higher and summer flows have been getting lower over time, but that these trends have been more variable in smaller catchments (Figure~\ref{figfour}, S1, S2). For example, winter maximum flows in small catchments ranged from nearly no change to dramatic 44\% increases per decade with decreases only as large as 15\%. Despite these extreme locations, on average the basin has only experienced moderate winter maximum flow increases of 5--10\%. Summer flows exhibited the opposite, with decreases in high flows over time. However, there was weak evidence of attenuation in the spring (Figure~\ref{figfour}, S1, S2). For example, in the spring the likelihood of network-driven attenuation of maximum flow trends was as low as 24\% (Figure~\ref{figfour}). The deterioration of the network's attenuation strength was coincident with the spring freshet, when snowmelt drives high flows across the basin. Synchronization events such as the freshet may subvert river network dampening mechanisms by homogenizing the region's response to climate shifts.

>>>>

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

\title{River networks dampen long-term hydrological signals of climate change}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Kyle A. Chezik}
\author[b,1]{Sean C. Anderson} 
\author[a,1]{Jonathan W. Moore}

\affil[a]{Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, 8888 University Dr., Burnaby, British Columbia V5A 1S6, Canada}
\affil[b]{School of Aquatic and Fishery Sciencies, University of Washington, Box 455020, Seattle, WA 98195, USA}

% Please give the surname of the lead author for the running footer
\leadauthor{Chezik}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Many ecosystems posses stabilizing processes that provide a natural defense against the impacts of climate change. The branching tree structure of river networks, provide a framework for aggregating the varied signals of climate change within a watershed and a possible mechanism for dampening long-term climate driven trends. Varied landscape attributes within a watershed produce a mosaic or portfolio of climate which in turn results in a variety of climate change manifestations. Here we demonstrate that large free-flowing river networks dampen local long-term flow trends by integrating a large and varied climate portfolio across a diverse and complex landscape. Integrating the benefits of these ecosystems into climate change mitigation stategies is essential in reducing the increasingly extreme impacts of climate change.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.W. Moore conceived the study and analysis, S.C. Anderson formalized the analysis, K.A. Chezik gathered and analyzed the data and wrote the paper. All authors contributed in all aspects to varying degrees.}
\authordeclaration{The authors declare no conflict of interest.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: kchezik@sfu.ca}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{rivers networks $|$ climate portfolio $|$ climate change $|$ trend attenuation $|$ climate dampening}

\begin{abstract}
	Climate change is dramatically altering temperate rivers --- snow-melt is shifting ever earlier \citep{Rauscher:2008]} floods are becoming more frequent \citep{Hirabayashi:2013}, and low flows more persistent \citep{McCabe:2014}. Here we describe a novel pathway by which river networks dampen local hydrologic climate change through the aggregation of upstream climate portfolio assets. We examine this hypothesis using flow and climate trend estimates (1970--2007), calculated at 55 hydrometric gauge stations and their contributing watersheds', within the vast Fraser River basin in British Columbia, Canada. Using a null hypothesis framework we compared our observed attenuation of river flow trends as a function of increasing area and climate trend diversity, with null-simulated estimates to gauge the likelihood and strength of our observations. We found the Fraser River dampened variability in downstream long-term discharge by >91\% across all flow trend response variables. For instance, median flow trends ranged from between -6 and 19\% per-decade among small watersheds, attenuating to between 0.8 and 1.6\% per-decade near the confluence. Our null-model approach indicated that dampening of annual median flow trends within the Fraser River was typically 3.3 times greater than would be expected of a watershed exhibiting a consistent basin-wide climate response. Although climate dampening strength varied seasonally, our findings indicate that large free-flowing rivers offer a powerful and largely unappreciated process of climate change mitigation. River networks that integrate a diverse climate portfolio can dampen local extremes and offer climate relief to riverine biota.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{A}s billions are spent adapting to climate change by such means as building dikes or improving stormwater drainage \citep{Narain:2011}, certain habitats and processes provide a natural defense system that mitigates climate change impacts \citep{Jones:2012}. For example, coastal habitats such as oyster reefs, mangrove forests, and eelgrass beds shield 67\% of the United States coastline and 1.4 million people from sea level rise and storms \citep{Arkema:2013}. Protection and enhancement of these natural systems has become a key component of many climate mitigation strategies \citep{Guerry:2015}, as they return multiple cost-effective services in contrast to engineered alternatives \citep{Jones:2012}. For instance, revitalization of the Mississippi delta not only improves storm protection, but also increases production and sustainability of fisheries and wildlife resources, protects the water supply, and reduces the impact of wastewater effluent \citep{Conservation:1998}. The value of climate buffering habitats stands only to increase as average global temperatures climb. The Paris Agreement of 2016 aims to limit global temperature rise to 1.5\degree{C} \citep{Hulme:2016}, a goal that concedes a minimum doubling of observed warming \citep{Hartmann:2013}; therefore, an urgent challenge and opportunity is discovering, maintaining and restoring systems that naturally buffer against the oncoming impacts of climate change.

In the face of changing global precipitation patterns \citep{Donat:2016}, river networks may offer an unappreciated defense against shifting flow regimes under climate change\citep{Hartmann:2013,Palmer:2009}. Earlier snowmelt \citep{Rauscher:2008}, reduced snow pack \citep{McCabe:2014}, shifts from snow to rain, and changes in the annual distribution of precipitation are all impacting stream flow \citep{Hartmann:2013}, resulting in flashier flows with increased frequency of flooding \citep{Hirabayashi:2013}, longer periods of drought and critically low flows for wildlife and agriculture \citep{Melillo:2014}. However, we hypothesize that large free-flowing river networks may naturally dampen these signals of climate change by integrating across varied landscapes and different manifestations of climate. Processes that aggregate across asynchronous components tend to dampen the aggregate, a process known as the portfolio effect \citep{Doak:1998}. Climate is expressed differently and asynchronously over the landscape \citep{George:2015} and on account of their branching architecture and directional flow, river networks integrate these varied climate manifestations \citep{Peterson:2013}, potentially dampening the local climatic response of sub-catchments. Large free-flowing rivers may have diverse climate portfolios, with sub-catchments acting as assets in their portfolio, thereby dampening local climate impacts. Portfolio effects in rivers are already known to dampen short-term variation \citep{Moore:2015,Yeakel:2014} but attenuation of long-term climate driven shifts have yet to be examined.

Here we consider whether river networks attenuate local long-term climate change through analysis of hydrometric data from one of the largest free-flowing rivers in North America. Located in British Columbia, Canada, the Fraser River drains an area approximately the size of the United Kingdom (\texttildelow 217,000 km\textsuperscript{2}) and in a rare combination is fairly well monitored while also having no dams on its main stem \citep{Vorosmarty:2010} (Fig.~\ref{fig:1}). This watershed drains interior high plateau, coastal mountains, and the Canadian Rockies, and thus integrates a diverse mosaic of landscapes, weather, and climate. As with other mid-latitude rivers \citep{Bindoff:2013}, the Fraser River has expressed increasing trends in discharge volume and variability over the last several decades \citep{Dery:2012,Morrison:2002}. The river's discharge is monitored by hydrometric gauge stations throughout the watershed. We analyzed 38 years of data collected concurrently at 55 of these stations and applied a novel analytical approach to examine whether this large river exhibits signals of long-term climate dampening.

\begin{figure}[h]
\centering
\includegraphics[width=17.8cm]{Fig1_Map.pdf}
	\caption{Climate trends in mean annual temperature (MAT) and mean annual precipitation (MAP) within the Fraser River basin overlayed on a digital elevation model of British Columbia, Canada. Flow gauge sites (dots) are scaled by the size of the contributing area. The map is projected in Albers BC UTMs to provide an equal area depiction of the region but labels are expressed in WGS84 latitude and longitude.}
\label{fig:1}
\end{figure}

\section*{Detecting Climate Change in the Fraser River Basin}
We focused on changes in hydrology as our riverine manifestation of climate signaling. As climate change alters precipitation patterns from snow to rain in temperate regions, rivers are generally predicted to have lower low-flows and higher high-flows, as well as earlier snowmelt \citep{Nijssen:2001}. These manifestations of rising temperatures are impacting species \citep{Xenopoulos:2006} and people \citep{Hirabayashi:2013}. Extracting annual and monthly metrics of flow from Fraser River basin gauge station records, we used rolling average maximum- and minimum-flow estimates to capture trends in extreme discharge and rolling average median-flow estimates to capture general annual and seasonal trends. Because the timing of flow is particularly important to the phenology of many organisms, we also calculated the day-of-year in which half the annual flow was reached.

To quantify changes in flow we fit generalized least squares models with an autoregressive error structure for each of our flow metrics between 1970 and 2007 for each site (\textit{Materials and Methods}). Long-term changes in flow metrics varied substantially across the watershed. For instance, annual maximum flow decreased in 40 sites but increased in 15 over the last four decades (-26 to 10\% change$\cdot$decade\textsuperscript{-1}, std. dev. = 6\%). Annual median flow showed similar variability but a greater tendency towards rising flow trends with 49 increasing and only six decreasing (-5 to 25\% change$\cdot$decade\textsuperscript{-1} std. dev. = 5\%). Thus, within the Fraser River basin, there are a variety of long-term trends in flow, suggesting a diversity of climate responses and landscape effects.

For a river network to act as a portfolio of climate and dampen the effects of climate change, expressions of climate and their trends over time need to vary throughout the network. Using the Western North America Climate tool (ClimateWNA), we quantified the diversity in climate trends within the Fraser River basin over the years concurrent with our hydrology trend estimates \citep{Wang:2016} (Fig.~\ref{fig:1}) (\textit{Materials and Methods}). We then summarized the climate portfolio of each flow gauge by calculating the standard deviation in climate trends within their catchments. Temperature and precipitation are likely to drive changing hydrology, therefore we included variables in our climate index that capture changes in extreme and mean temperature, as well as changes in precipitation volume and physical state. The standard deviations of these climate variables were scaled between 0 and 1 and summed into a general climate index for each catchment. Following our hypothesis, this climate portfolio index increases with area (Fig.~\ref{fig:2}), rapidly asymptoting in catchments greater than 40,000 km\textsuperscript{2}.

\begin{figure}%[tbhp]
\centering
\includegraphics[width=.8\linewidth]{Fig_ClimPort.pdf}
	\caption{Climate portfolio index as a function of catchment area. The mean trend line (blue) was estimated using robust linear regression and standard error estimates (grey) were calculated using the weighted standard deviation estimate.}
\label{fig:2}
\end{figure}

\section*{Quantifying the River Network Portfolio Effect}
Climate portfolio diversity is largely a function of area (Fig.~\ref{fig:2}). Larger areas contain greater landscape complexity and thus a greater variety of climatic expressions. The influence of any individual sub-basin on downstream flow trends is also a function of area. Larger catchments simply aggregate more water such that their climate portfolio will carry more weight on downstream dampening. The interplay between climate portfolio and catchment area on the impact of downstream flow trend dampening makes area a more robust response variable than simply the climate portfolio index. Therefore, to test whether networks dampen climate signals we regressed site-specific trend estimates (e.g., \% change$\cdot$decade\textsuperscript{-1}) onto each site's catchment area using a generalized least squares model (\textit{Materials and Methods}). To quantify attenuation, we modeled the change in trend variability with the change in catchment area using an exponential variance function \citep{Pinheiro:2006}. The variance function allowed us to predict the range of flow trend values that would be expected as watershed area increased. A decreasing range of flow trends with increasing watershed area indicates a dampening network effect.

In the observed data, small watersheds exhibited greater variability around their trend estimates than large watersheds, likely because of greater short-term variation in small catchments \ref{fig:3} \citep{Moore:2015}. This relationship between watershed size and trend certainty may pull small watershed trend estimates away from zero, thereby creating the appearance of decreasing trend variability among sites as watershed size increases. Using a null-hypothesis framework (\textit{Materials and Methods}) we simulated time-series for each site with no underlying trend. These simulations were generated by a random walk process with the observed standard deviation and autocorrelation parameters ($\hat{\upsigma}$, $\hat{\upphi}$) at our sites. Using the same generalized least squares model applied to the observed data, we estimated trends for 1000 simulated time series at each site for each response variable (e.g., Figure S3). This simulation process, a form of parametric bootstrapping, created distributions of null-expectations with which we compared our observed results.

\begin{figure}[h]
\centering
\includegraphics[width=17.8]{Fig_AnnFun.pdf}
	\caption{Annual flow trend attenuation within the Fraser River basin. (\textbf{Left}) Trend estimates $\pm$ one standard error (SE, grey) plotted against watershed area (km\textsuperscript{2}), colored by climate portfolio strength (green = small, blue = large). Blue lines represent observed attenuation; orange and red represent simulated attenuation that is weaker and stronger than observed. (\textbf{Right}) Density plots show null model simulated variance exponents ($\hat{\updelta}$) and the proportion on either side of the observed variance exponent (blue). Flow metrics include long-term flow-timing shifts (change per decade in day-of-year (DOY) to half annual flow), where decreasing trends suggest more annual flow is occurring earlier in the year (a), and the percent change per decade in minimum (b), maximum (c) and median (d) annual flow. Simulated lines ignore variance in the intercept and slope to focus visually on attenuation.}
\label{fig:3}
\end{figure}

\section*{Results}
\subsection{Climate Portfolios and Flow Trend Dampening.} Long-term changes in hydrology were less variable in larger catchments, as predicted, demonstrating between 92 and 96\% dampening of flow trends moving from headwater catchments to the network's confluence. In other words, the largest catchments were approximately 10 times more stable in their climate response than the smallest catchments. For instance, trends ranged between a 6\% reduction and 19\% increase in median flow per decade among small watersheds, while large watersheds ranged between 0.8 and 1.6\% change per decade. Similar attenuation was seen in flow timing with a 94\% reduction in the day-of-year to half-annual-flow trends such that small watersheds were reaching half their annual flow between 8 days later and 25 days earlier while large watersheds have only shifted 2 to 4 days earlier between 1970 and 2007.  
Our null model approach illustrates that support for a river network portfolio effect is greater than 90\% for all four annual flow trend metrics and as high as 98\% for median-annual-flow trends (Fig.~\ref{fig:3}). Trend attenuation was 3.0 (0.7--9.4 this and hereafter are 90\% intervals), 4.8 (1.1--13.8), 3.4 (0.9--9.0) and 5.3(1.4--14.8) times greater than the null model for annual-flow timing, and minimum, maximum and median annual-flow respectively. These statistics provide strong support for the hypothesis that river network portfolio effects contribute to climate dampening in large rivers.

\subsection{Seasonally Shifting Flow Trend Dampening}
Regional flow-trends and attenuation strength varied by season. Analysis of monthly trends revealed that winter flows are getting higher and summer flows are getting lower over time, but that these trends have been more variable in smaller catchments (Fig.~\ref{fig:4}, \textit{SI Appendix}, Figs. S~\ref{fig:S1}, S~\ref{fig:S2}). For example, winter maximum flows in small catchments ranged from nearly no change to dramatic 47\% increases per decade with decreases only as large as 9\%. Despite these extreme locations, on average the basin has only experienced moderate winter maximum flow increases of 5-9\%. Summer flows exhibited the opposite, with decreases in high flows over time. However, there was weak evidence of attenuation in the spring (Fig.~\ref{fig:4}, \textit{SI Appendix},Figs. S~\ref{fig:S1}, S~\ref{fig:S2}). Overall, climate portfolios tended to be greater during stable periods of the year and decrease during seasonal transitions (\textit{SI Appendix}, Fig.~\ref{S7}). In the spring the likelihood of network-driven attenuation of maximum flow trends was as low as 25\% (Fig.~\ref{fig:4}). The deterioration of the network's attenuation strength was coincident with the spring freshet, when snowmelt drives high flows across the basin. Synchronization events such as the freshet may subvert river network dampening mechanisms by homogenizing the region's response to seasonally driven climate shifts.

\begin{figure}%[tbhp]
\centering
\includegraphics[width=.8\linewidth]{Figure4.pdf}
	\caption{Monthly maximum flow trend attenuation within the Fraser River basin. (\textbf{Left}) Fraser River's basin-wide maximum-flow trend estimates (i.e., intercept = vertical grey lines) by month with density distributions of null-model simulations. Observed values falling further from the center of the density distribution suggest greater evidence for changes in maximum flow and a greater shift in magnitude. (\textbf{Center}) Observed monthly Fraser River maximum-flow variance exponent ($\hat{\updelta}$, blue) and associated density distribution of simulated $\hat{\updelta}$ estimates. Decimal values represent the percent of simulated data exhibiting weaker attenuation (yellow) than observed. (\textbf{Right}) Trend estimates $\pm$ one standard error (SE, grey) plotted against watershed area (km\textsuperscript{2}), colored by climate portfolio strength (green = small, blue = large), for four seasonally representative months. These reflect months in the prior columns and describe the variation in percent change per decade of maximum flow among sites. Simulated lines ignore variance in the intercept and slope to focus visually on attenuation.}
\label{fig:4}
\end{figure}
